---
title: "Stratified 10 year risk of all cause dementia and dementia subtybes - cohorts"
author: "JQT"
date: "26/01/2022"
output:
  html_document: default
  pdf_document: default
---

Please change cohorts, author and date in the above.

# Description

To use this script two files must be prepared - one with cohort phenotype information and one with genotype information. <br><br>

The endpoints are defined as: <br>
  alz: Alzheimer's disease (ICD10: F00, G30 or similar) <br>
  vas: vascular dementia (ICD10: F01 or similar) <br>
  vard: vascular related dementia (ICD10: F01, F03 or similar) and no alz diagnosis (ICD10: F00, G30) <br>
  alldem: all cause dementia (ICD10: F00, F01, F03 G30 or similar) <br><br>

If other endpoint definition is used please add definition here in Rnotebook.<br><br>

Endpoint-exit dates are defined as the earliest date of event, death, study end, loss to followup. <br>
Four endpoint exit dates are used: <br>
  alz_date: the earliest date of alz, death, study end, loss to followup, no censoring of other dementia diagnosis <br>
  vas_date: the earliest date of vas, death, study end, loss to followup, no censoring of other dementia diagnosis <br>
  vard_date: the earliest date of vard, death, study end, loss to followup or alz-diagnosis date <br>
  alldem_date: the earliest date of alldem, death, study end, loss to followup <br><br>

The file with phenotype information must include: <br>
  subjid : an subject id for merging phenotype and genotype files. <br>
  age: age at baseline <br>
  sex: sex coded as 0:women, 1:men <br>
  entrydate: entry date <br>
  died: if individual died, 0:alive, 1: died <br>
  dm: diabetes status, 0: no diabetes, 1: diabetes diagnosis <br>
  smoker: ever-never smoker, 0: never smoked at baseline, 1: had/was smoker at baseline <br>
  education: education level, 0: >8 years of schooling, 1: <=8 years of schooling <br>
  alz: Alzheimer's disease coded as 0: no disease, 1: Alzheimer's disease <br>
  alz_date: alz endpoint date <br>
  vas: vascular dementia coded as 0: no disease, 1: vascular dementia <br>
  vas_date: vas endpoint date <br>
  vard: vascular related dementia coded as 0: no disease, 1: vascular related dementia <br>
  vard_date: vas endpoint date <br>
  alldem: all cause dementia coded as 0: no disease, 1: all cause dementia <br>
  alldem_date: alldem endpoint date <br> 
  apoe: APOE-genotype codes as: e22,e32,e33,e42,e43,e44 <br><br>

The file with genotype information should contain either genotyped or imputed dosages of the SNPs in the models. <br>
Se the supplied csv-files defining the models. <br>
Two ploygenic scores are used in the analysis. First is based on IGAP2013 paper and the second on EADB2022 paper. <br><br>

If genotypes are provided as a genotyped file is must include: <br>
  id: an subject id for merging phenotype and genotype files. <br>
  rsid-name: Coded with genotype call e.g. GA, CT <br><br>
!! Contact Jesper for script to use genotypes. This script uses vcf-imputed data.

If genotypes are supplied as imputed dosages then please provide a vcf file with the dosages and an info-file with the imptation quality and MAF. <br>
(Supplied a script to compile VCF-file with SNPs and to extract information from info files) <br>


```{r setup, include=FALSE}

rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(data.table)
library(vcfR)
library(rmarkdown)
library(prodlim)
library(pec)
library(ggplot2)
library(riskRegression)
library(foreign)
library(moments)
library(car)
library(survival)


# Set working directory
knitr::opts_knit$set(root.dir = 'L:\\LovbeskyttetMapper\\Forskningsdrev\\Jesper\\EADB\\10_year_risk_prediction\\Rscripts AD10yearrisk\\')

# Set your cohort name for plots and results
cohortname <- "CGPS exome"
```

# Prepare data

## Read  VCF-data
```{r read datafiles}

# Read VCF-file
print('Excg')
geno_imp <- read.vcfR("./data/Excg_EADB2022SNPS.dose.vcf",)


```


```{r extract dosage information from VCF-file}

# Extract FIX and dosages from vcfR-object in useful dataframes
geno_imp.FIX <- as.data.frame(getFIX(geno_imp))
geno_imp.DS <- as.data.frame(t(extract.gt(geno_imp, element="DS")))

#-------------------------------------------------------------------------------
# Extract sample-id from VCF_file-sample-id where name has been doubled.
# Ex-CGPS
geno_imp.rn <- row.names(geno_imp.DS)
rn <- t(as.data.frame(gregexpr("_", geno_imp.rn)))
geno_imp.rn <- as.data.frame(geno_imp.rn)
colnames(geno_imp.rn)[1] <- "rownam"
geno_imp.rn <- cbind(geno_imp.rn,rn)

geno_imp.rn <- geno_imp.rn %>% mutate(sample_id = substr(rownam,0,rn -1) )
sample_id <- geno_imp.rn$sample_id
geno_imp.DS <- cbind(sample_id,geno_imp.DS)


rm(rn,geno_imp.rn,geno_imp,sample_id)

```
## Read infofile data
```{r read infofiles}
geno_info <- fread("./data/ExCG_ImputationInfo_20220216.txt",dec = ",")

```

```{r find snps with good imputation quality}
geno_info_QC <- geno_info[geno_info$Rsq >= 0.4,]

geno_snpid <- c("sample_id",geno_info_QC$LOCID)
geno_imp.DS_QC <- geno_imp.DS %>% select(all_of(geno_snpid)) 

```

## EADB2022 model

```{r EADB 2022 model - Excg}
# read gene score model EADB 2022
gsmodel2022 <- fread("./data/Allele_score_EADB2022.csv",dec = ",")

gsmodel2022 <- gsmodel2022 %>% mutate(snpid = paste0(chr,":",`position(GRCH37)`))
gsmodel2022$OR <- as.numeric(gsmodel2022$OR)
gsmodel2022$maf_cohort <- "NA"
gsmodel2022$note <- "NA"

# for loop QC-snps in model
for (i in 2:length(geno_snpid) ){
# Check SNPmodel for increasing risk variant (OR<>1)
#i=2
SNP_beta <-  log(gsmodel2022$OR[gsmodel2022$snpid==geno_snpid[i]] )

if (SNP_beta >= 0)
  {
    increasing_allele = 1
  } else {
    increasing_allele = -1  
  }

# Check for ref/alt match in data and model
model_ref <-  gsmodel2022$ref0[gsmodel2022$snpid==geno_snpid[i]] 
model_alt <-  gsmodel2022$alt[gsmodel2022$snpid==geno_snpid[i]] 

data_ref <- geno_imp.FIX$REF[geno_imp.FIX$ID==geno_snpid[i]]
data_alt <- geno_imp.FIX$ALT[geno_imp.FIX$ID==geno_snpid[i]]

if (model_ref == data_ref)
 { 
  if (model_alt == data_alt)
   {
     # reference and alternate allele match
     model_data_direction = 1
   } else {
     # reference allele match, but alternate alleles differ - note inserted in model_table
     gsmodel2022$note[gsmodel2022$snpid==geno_snpid[i]] = "alt differ"
     next
   }
 } else {
   if (model_ref == data_alt)
   {
     if (model_alt == data_ref)
      {
        model_data_direction = -1
      } else {
        gsmodel2022$note[gsmodel2022$snpid==geno_snpid[i]] = "model_alt no match"
        next
      }
   } else {
     gsmodel2022$note[gsmodel2022$snpid==geno_snpid[i]] = "model_ref no match"
     next
   }
 }
# copy data-maf to model_table
gsmodel2022$maf_cohort[gsmodel2022$snpid==geno_snpid[i]] <- geno_info$ALT_AF[geno_info$LOCID==geno_snpid[i]]




# Combine increasing allele and ref/alt match
dosage_change <- as.numeric(increasing_allele * model_data_direction)

# Multiply abs(log(OR)) to dosage - new column
# if direction opposite or then dosage(R)=2-dosage(A)

geno_imp.DS_QC[,all_of(geno_snpid[i])] <- as.numeric(geno_imp.DS_QC[,all_of(geno_snpid[i])])

new_column <- paste0(geno_snpid[i],"_wds")

temp_ds <- test <- geno_imp.DS_QC %>% select(all_of(geno_snpid[i]))
#temp_ds[,1] <- as.numeric(temp_ds[,1])                                            

temp_ds <- temp_ds %>% transmute({{new_column}} := case_when(
                dosage_change == 1  ~ temp_ds[,1] * abs(SNP_beta),
                dosage_change == -1 ~ (2-temp_ds[,1]) * abs(SNP_beta)
)
)

geno_imp.DS_QC <- cbind(geno_imp.DS_QC,temp_ds)


# end forloop
}
# rowsums(effect*dosage) - new column

geno_snpid_wds <- unlist(lapply(geno_snpid[2:length(geno_snpid)],function(x){paste0(x,"_wds")}))


geno_imp.DS_QC$was2022 <- rowSums(geno_imp.DS_QC[,all_of(geno_snpid_wds)] )

# histogram was2022
hist(geno_imp.DS_QC$was2022,breaks=20)
# Test for normality
qqPlot(geno_imp.DS_QC$was2022)
skewness(geno_imp.DS_QC$was2022)
kurtosis(geno_imp.DS_QC$was2022)

#Summary
summary(geno_imp.DS_QC)

# Create WAS groups

geno_imp.DS_QC <- geno_imp.DS_QC %>% mutate(was2022_gr =  ntile(was2022, 4))
geno_imp.DS_QC$was2022_gr <- as.factor(geno_imp.DS_QC$was2022_gr)
summary(geno_imp.DS_QC$was2022_gr)


rm(geno_info,geno_info_QC,geno_snpid,geno_snpid_wds,test,temp_ds)
rm(data_alt,data_ref,dosage_change,i,increasing_allele,model_alt,model_ref,model_data_direction,new_column,SNP_beta)
```






## Read phenotypes - EADB2022

```{r read phenotypes}

pheno <- fread("./data/EADB2022_phenotypes_cgps_20220208.csv")

```

```{r change column-names - phenotypes}

# If needed change column names or remove columns to fit with code-names.
# Otherwise comment out.
pheno <- subset(pheno,select=-c(sample_id,populid.y))
pheno <- pheno %>% rename(sample_id=exome_Sample_ID)

pheno <- pheno %>% rename(died=dod)
pheno <- pheno %>% rename(smoker=smoke)

pheno <- pheno %>% rename(alz=alzheimer)
pheno <- pheno %>% rename(vas=vascular_dementia)
pheno <- pheno %>% rename(alldem=demensall)

pheno <- pheno %>% rename(alz_date=alzheimerdato)
pheno <- pheno %>% rename(vas_date=vascular_dementiadato)
pheno <- pheno %>% rename(alldem_date=demensalldato)


```

## Prepare phenotypes
```{r prepare phenotypes - CCHS}

# APOE genotype
print("APOE")
pheno$apoe <- as.factor(as.character(pheno$apoe))
pheno$apoe <- relevel(pheno$apoe,ref="e33")

summary(pheno$apoe)

# Sex
print("Sex")
pheno$sex<- as.factor(as.character(pheno$sex))
summary(pheno$sex)

#Age groups
print("age_gr")
pheno <- pheno %>% mutate(age_gr = cut(age, 
                    breaks = c(19, 50, 60, 70, 80, Inf),
                    labels = c("20-50", "50-60", "60-70", "70-80", ">80"))
                )
summary(pheno$age_gr)

# diabetes
print("dm")
pheno$dm <- as.factor(as.character(pheno$dm))
summary(pheno$dm)

# smoker
print("smoker")
pheno$smoker <- as.factor(as.character(pheno$smoker))
summary(pheno$smoker)

# educate
print("educate")
pheno$educate <- as.factor(as.character(pheno$educate))
summary(pheno$educate)

```

## Prepare endpoints

```{r prepare endpoints and related}

# Alzheimer
print("alz")
pheno$alz <- as.factor(as.character(pheno$alz))
summary(pheno$alz)

# Vascular
print("vas")
pheno$vas <- as.factor(as.character(pheno$vas))
summary(pheno$vas)

# vascular related dementia
print("vard")
pheno$vard <- as.factor(as.character(pheno$vard))
summary(pheno$vard)

# All dementia
print("alldem")
pheno$alldem <- as.factor(as.character(pheno$alldem))
summary(pheno$alldem)

# Death
print("died")
pheno$died <- as.factor(as.character(pheno$died))
summary(pheno$died)



```
## Prepare time-variables
```{r Prepare time-variables}

hist(pheno$entrydate,breaks="years",freq=TRUE,xlab="examination date")
hist(pheno$age,freq=TRUE,,xlab="age")
hist(pheno$alz_date[pheno$alz==1],breaks="years",freq=TRUE,xlab="Alzheimer disease date")
hist(pheno$vas_date[pheno$vas==1],breaks="years",freq=TRUE,xlab="Vascular dementia date")
hist(pheno$vard_date[pheno$vard==1],breaks="years",freq=TRUE,xlab="Vascular related dementia date")
hist(pheno$alldem_date[pheno$alldem==1],breaks="years",freq=TRUE,xlab="All dementia date")
```

```{r prepare follow-up time}
pheno <- pheno %>% mutate( futime_alz = as.numeric((alz_date - entrydate)) )
summary(pheno$futime_alz)

hist(pheno$futime_alz,breaks=25,freq=TRUE,xlab="Followup time Alzheimer disease endpoint")

pheno <- pheno %>% mutate( futime_vas = as.numeric((vas_date - entrydate)) )
summary(pheno$futime_vas)

hist(pheno$futime_vas,breaks=25,freq=TRUE,xlab="Followup time Vascular dementia endpoint")

pheno <- pheno %>% mutate( futime_vard = as.numeric((vard_date - entrydate)) )
summary(pheno$futime_vard)

hist(pheno$futime_vard,breaks=25,freq=TRUE,xlab="Followup time vascular related dementia endpoint")

pheno <- pheno %>% mutate( futime_all = as.numeric((alldem_date - entrydate)) )
summary(pheno$futime_all)

hist(pheno$futime_all,breaks=25,freq=TRUE,xlab="Followup time All dementia endpoint")

# Check and exclude individuals with dementia at baseline

exclude_all <- pheno[,alldem_date < entrydate]
summary(exclude_all)
pheno <- pheno[!exclude_all,]

rm(exclude_alz,exclude_all)
```

# 10 year risk analysis

## Set-up analysis

```{r Join tabels Excc}
geno_df <-  geno_imp.DS_QC[,c('sample_id','was2022_gr')]
pheno_df <- pheno[,c('sample_id','sex','age_gr','apoe','dm','smoker','educate','futime_alz','alz','futime_vas','vas','futime_vard','vard','futime_all','alldem','died')]

df <- merge(pheno_df,geno_df,by.x="sample_id", by.y='sample_id')
rm(geno_df,pheno_df)
# Summary entire dataset with apoegenotype nulls
summary(df)

# Table age_gr, alz
print("alz")
table(df$age_gr,df$alz)
# Table age_gr, vas
print("vas")
table(df$age_gr,df$vas)
# Table age_gr, nonalz
print("vard")
table(df$age_gr,df$vard)
# Table age_gr, all dementia
print("alldem")
table(df$age_gr,df$alldem)

# apoegenotype, alz
print("alz")
table(df$apoe,df$alz)
# apoegenotype, vas
print("vas")
table(df$apoe,df$vas)
# apoegenotype, nonalz
print("vard")
table(df$apoe,df$vard)
# apoegenotype, all dementia
print("alldem")
table(df$apoe,df$alldem)
```

```{r Tabular individuals in groups}
# sex, age, allele_group, APOE genotype
# diabetes, smoking, education

# Weighted allele score

tabular_df <- as.data.frame(table(df$apoe,df$was2022_gr,df$sex,df$age_gr,df$dm,df$smoker,df$educate))
colnames(tabular_df) <- c("APOE","was_gr","sex","age_gr","dm","smoker","educate","count")

filename <- "./Results/tabular_was_df_EADB2022_"
write.table(tabular_df, paste0(filename,cohortname,".txt"), append = FALSE, sep = " ",row.names = F)

#tabular_df2
```


```{r Cause specific Cox models}

# Cox regression model for Alzheimer's disease 
alz_cox_was <- coxph(Surv(time=futime_alz,event=alz)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, id=sample_id)
summary(alz_cox_was)

# Cox regression  model for Vascular dementia
vas_cox_was <- coxph(Surv(time=futime_vas,event=vas)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, id=sample_id)
summary(vas_cox_was)

# Cox regression  model for Non-AD 
vard_cox_was <- coxph(Surv(time=futime_vard,event=vard)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, id=sample_id)
summary(vard_cox_was)

# Cox regression  model for all dementia
alldem_cox_was <- coxph(Surv(time=futime_all,event=alldem)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, id=sample_id)
summary(alldem_cox_was)


```




## Prepare compeeting risk variables 
```{r prepare compeeting risk variables}
# Alzheimer compeeting risk
df <- df %>% mutate(alz_comp = case_when(
                  alz==1 ~ "2",
                  died == 1 &  alz==0 ~ "1",
                  alz==0 & died==0 ~ "0")
                )
df$alz_comp <- as.factor(as.character(df$alz_comp))
print("alz")
summary(df$alz_comp)

# Vascular compeeting risk
df <- df %>% mutate(vas_comp = case_when(
                  vas==1 ~ "2",
                  died == 1 &  vas==0 ~ "1",
                  vas==0 & died==0 ~ "0")
                )
df$vas_comp <- as.factor(as.character(df$vas_comp))
print("vas")
summary(df$vas_comp)

# non_ad compeeting risk
df <- df %>% mutate(vard_comp = case_when(
                  vard==1 ~ "2",
                  died == 1 &  vard==0 ~ "1",
                  vard==0 & died==0 ~ "0")
                )
df$vard_comp <- as.factor(as.character(df$vard_comp))
print("vard")
summary(df$vard_comp)

# All dementia compeeting risk
df <- df %>% mutate(alldem_comp = case_when(
                  alldem==1 ~ "2",
                  died == 1 &  alldem==0 ~ "1",
                  alldem==0 & died==0 ~ "0")
                )
df$alldem_comp <- as.factor(as.character(df$alldem_comp))
print("alldem")
summary(df$alldem_comp)

```

## Fine-Grey risk regression models
```{r Fine-Grey models}

# Fine-Grey risk regression regression model for Alzheimer's disease 
alz_fgr_was <- FGR(Hist(time=futime_alz,event=alz_comp)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, cause=2)
summary(alz_fgr_was)

# Fine-Grey risk regression regression model for Vascular dementia
vas_fgr_was <- FGR(Hist(time=futime_vas,event=vas_comp)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, cause=2)
summary(vas_fgr_was)

# Fine-Grey risk regression regression model for Non-AD 
vard_fgr_was <- FGR(Hist(time=futime_vard,event=vard_comp)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, cause=2)
summary(vard_fgr_was)

# Fine-Grey risk regression regression model for all dementia
alldem_fgr_was <- FGR(Hist(time=futime_all,event=alldem_comp)~sex+age_gr+apoe+was2022_gr+dm+smoker+educate, data=df, cause=2)
summary(alldem_fgr_was)
```

## 10 year Risk estimates

```{r setup and compute 10 year risk estimates}
# Setup array for results

# loop - sex
# loop - age_group
# loop - APOE_genotype
# loop - Allele_score
# loop - Diabetes status
# loop - Smoking_n_e
# loop - Education_gr

sex <- c(0,1)
age_gr <- c(0,1,2,3,4)
apoe <- relevel(as.factor(c("e33","e22","e32","e42","e43","e44")),ref="e33")
was2022_gr <- c(0,1,2,3)
dm <- c(0,1)
smoker <- c(0,1)
educate <- c(0,1)

m1 <- CJ(sex,age_gr,apoe,was2022_gr,dm,smoker,educate)

m1$sex <- as.factor(m1$sex)
m1$age_gr <- as.factor(m1$age_gr)
m1$was2022_gr <- as.factor(m1$was2022_gr)
m1$dm <- as.factor(m1$dm)
m1$smoker<- as.factor(m1$smokerr)
m1$educate<- as.factor(m1$educate)

rm(sex,age_gr,apoe,was2022_gr,dm,smoker,educate)

# Risk prediction times 5,10,15,20 years

pred_alz <- as.data.frame(predictEventProb(alz_fgr_was,times=c(5*365.24,10*365.24,15*365.24,20*365.24),newdata=m1,cause=2))
colnames(pred_alz) <- c("alz_pred_05","alz_pred_10","alz_pred_15","alz_pred_20")

pred_vas <- as.data.frame(predictEventProb(vas_fgr_was,times=c(5*365.24,10*365.24,15*365.24,20*365.24),newdata=m1,cause=2))
colnames(pred_vas) <- c("vas_pred_05","vas_pred_10","vas_pred_15","vas_pred_20")

pred_vard <- as.data.frame(predictEventProb(vard_fgr_was,times=c(5*365.24,10*365.24,15*365.24,20*365.24),newdata=m1,cause=2))
colnames(pred_vard) <- c("vard_pred_05","vard_pred_10","vard_pred_15","vard_pred_20")

pred_alldem <- as.data.frame(predictEventProb(alldem_fgr_was,times=c(5*365.24,10*365.24,15*365.24,20*365.24),newdata=m1,cause=2))
colnames(pred_alldem) <- c("alldem_pred_05","alldem_pred_10","alldem_pred_15","alldem_pred_20")



alz_df <- cbind(m1,pred_alz)
vas_df <- cbind(m1,pred_vas)
vard_df <- cbind(m1,pred_vard)
alldem_df <- cbind(m1,pred_alldem)

rm(pred_alz,pred_vas,pred_vard,pred_alldem,m1)

```

```{r setup variables for risk plots}
# Set up variables for risk plots

age_gr.labs <- c("50-59", "60-69", "70-79",">80")
names(age_gr.labs) <- c("1", "2", "3","4")

smoker.labs <- c("No smoking", "Smoking")
names(smoker.labs) <- c("0", "1")

educate.labs <- c("Education <8y", "Education >8y")
names(educate.labs) <- c("1", "0")

dm.labs <- c("No diabetes", "Diabetes")
names(dm.labs) <- c("0", "1")

cols <- c("0"="#00994C","1"="#66CC00","2"="#FFFF33" ,"3"="#FF9933", "4"="#FF0000","5"="#CC0000","6"="#994C00")


```

```{r Alz risk charts}

# Alzheimer's disease risk charts

alz_df <- alz_df %>% mutate(apoe_s = case_when(
                  apoe == "e22" ~ 1,
                  apoe == "e32" ~ 2,
                  apoe == "e33" ~ 3,
                  apoe == "e42" ~ 4,
                  apoe == "e43" ~ 5,
                  apoe == "e44" ~ 6)
                )

alz_df$apoe_s<- as.factor(alz_df$apoe_s)

alz_df <- alz_df %>% mutate(col = case_when(
                  alz_pred_10 < 0.01 ~ "0",
                  alz_pred_10 >= 0.01 & alz_pred_10 < 0.05 ~ "1",
                  alz_pred_10 >= 0.05 & alz_pred_10 < 0.10 ~ "2",
                  alz_pred_10 >= 0.10 & alz_pred_10 < 0.15 ~ "3",
                  alz_pred_10 >= 0.15 & alz_pred_10 < 0.20 ~ "4",
                  alz_pred_10 >= 0.20 & alz_pred_10 < 0.30 ~ "5",
                  alz_pred_10 >= 0.30 ~ "6")
                )

# order age_gr for plot 
summary(alz_df$age_gr)
alz_df$age_gr <- factor(alz_df$age_gr, levels = c(4, 3, 2, 1,0))


for (i in 0:1) {
  for (j in 0:1) {
  
pd_alz <- alz_df[sex==i & educate==j & age_gr != 0,]

titlename <- paste0(cohortname," - Alzheimer 10 year absolute risk - sex=" ,i, " edu=" ,j)

ggplot(pd_alz,aes(x = apoe_s, y = was2022_gr, fill = col))  +
  geom_tile() +
  geom_text(aes(label = lapply(alz_pred_10*100.0, as.integer)),fontface = "bold") +
  scale_fill_manual(values=cols)+
  labs(x="APOE genotype", y="Allele score") + 
  scale_x_discrete( labels=c(expression(paste(epsilon,"22")),expression(paste(epsilon,"32")),expression(paste(epsilon,"33")),expression(paste(epsilon,"42")),expression(paste(epsilon,"43")),expression(paste(epsilon,"44")))) + 
  scale_y_discrete( labels=c("0-25%","25-50%","50-75%","75-100%")) + 
  theme(axis.ticks = element_blank(), panel.background = element_blank(),legend.position = "none",      axis.title=element_text(size=12,face="bold")) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0 , b = 0, l = 0))) +
  theme(axis.text=element_text(size=8,face="italic"), strip.background = element_blank()) +
  facet_grid(age_gr ~ dm + smoker , labeller = labeller(age_gr = age_gr.labs, smoker = smoker.labs, dm = dm.labs)) +
  theme(strip.text.x = element_text( size = 10, face = "bold"),
        strip.text.y = element_text( size = 10, face = "bold")) +
  ggtitle(titlename) +
  coord_fixed()


plotname <- paste0("./Results/",cohortname,"_EADB2022_Riskplot_10y_alz_sex",i,"_edu",j,".pdf")

ggsave(plotname,device="pdf",width=20,dpi=300,units="cm")

}
}

```

```{r vas risk charts}
# Vascular dementia risk charts

vas_df <- vas_df %>% mutate(apoe_s = case_when(
                  apoe == "e22" ~ 1,
                  apoe == "e32" ~ 2,
                  apoe == "e33" ~ 3,
                  apoe == "e42" ~ 4,
                  apoe == "e43" ~ 5,
                  apoe == "e44" ~ 6)
                )

vas_df$apoe_s<- as.factor(vas_df$apoe_s)

vas_df <- vas_df %>% mutate(col = case_when(
                  vas_pred_10 < 0.01 ~ "0",
                  vas_pred_10 >= 0.01 & vas_pred_10 < 0.05 ~ "1",
                  vas_pred_10 >= 0.05 & vas_pred_10 < 0.10 ~ "2",
                  vas_pred_10 >= 0.10 & vas_pred_10 < 0.15 ~ "3",
                  vas_pred_10 >= 0.15 & vas_pred_10 < 0.20 ~ "4",
                  vas_pred_10 >= 0.20 & vas_pred_10 < 0.30 ~ "5",
                  vas_pred_10 >= 0.30 ~ "6")
                )

# order age_gr for plot 
summary(vas_df$age_gr)
vas_df$age_gr <- factor(vas_df$age_gr, levels = c(4, 3, 2, 1,0))


for (i in 0:1) {
  for (j in 0:1) {
  
pd_vas <- vas_df[sex==i & educate==j & age_gr != 0,]

titlename <- paste0(cohortname," - Vascular dementia 10 year absolute risk - sex=" ,i, " edu=" ,j)

ggplot(pd_vas,aes(x = apoe_s, y = was2022_gr, fill = col))  +
  geom_tile() +
  geom_text(aes(label = lapply(vas_pred_10*100.0, as.integer)),fontface = "bold") +
  scale_fill_manual(values=cols)+
  labs(x="APOE genotype", y="Allele score") + 
  scale_x_discrete( labels=c(expression(paste(epsilon,"22")),expression(paste(epsilon,"32")),expression(paste(epsilon,"33")),expression(paste(epsilon,"42")),expression(paste(epsilon,"43")),expression(paste(epsilon,"44")))) + 
  scale_y_discrete( labels=c("0-25%","25-50%","50-75%","75-100%")) + 
  theme(axis.ticks = element_blank(), panel.background = element_blank(),legend.position = "none",      axis.title=element_text(size=12,face="bold")) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0 , b = 0, l = 0))) +
  theme(axis.text=element_text(size=8,face="italic"), strip.background = element_blank()) +
  facet_grid(age_gr ~ dm + smoker , labeller = labeller(age_gr = age_gr.labs, smoker = smoker.labs, dm = dm.labs)) +
  theme(strip.text.x = element_text( size = 10, face = "bold"),
        strip.text.y = element_text( size = 10, face = "bold")) +
  ggtitle(titlename) +
  coord_fixed()


plotname <- paste0("./Results/",cohortname,"_EADB2022_Riskplot_10y_vas_sex",i,"_edu",j,".pdf")

ggsave(plotname,device="pdf",width=20,dpi=300,units="cm")

}
}

```

```{r vard risk charts}
# Vascular related disease risk charts


vard_df <- vard_df %>% mutate(apoe_s = case_when(
                  apoe == "e22" ~ 1,
                  apoe == "e32" ~ 2,
                  apoe == "e33" ~ 3,
                  apoe == "e42" ~ 4,
                  apoe == "e43" ~ 5,
                  apoe == "e44" ~ 6)
                )

vard_df$apoe_s<- as.factor(vard_df$apoe_s)

vard_df <- vard_df %>% mutate(col = case_when(
                  vard_pred_10 < 0.01 ~ "0",
                  vard_pred_10 >= 0.01 & vard_pred_10 < 0.05 ~ "1",
                  vard_pred_10 >= 0.05 & vard_pred_10 < 0.10 ~ "2",
                  vard_pred_10 >= 0.10 & vard_pred_10 < 0.15 ~ "3",
                  vard_pred_10 >= 0.15 & vard_pred_10 < 0.20 ~ "4",
                  vard_pred_10 >= 0.20 & vard_pred_10 < 0.30 ~ "5",
                  vard_pred_10 >= 0.30 ~ "6")
                )

# order age_gr for plot 
summary(vard_df$age_gr)
vard_df$age_gr <- factor(vard_df$age_gr, levels = c(4, 3, 2, 1,0))


for (i in 0:1) {
  for (j in 0:1) {
  
pd_vard <- vard_df[sex==i & educate==j & age_gr != 0,]

titlename <- paste0(cohortname," - vascular related dementia 10 year absolute risk - sex=" ,i, " edu=" ,j)

ggplot(pd_vard,aes(x = apoe_s, y = was2022_gr, fill = col))  +
  geom_tile() +
  geom_text(aes(label = lapply(vard_pred_10*100.0, as.integer)),fontface = "bold") +
  scale_fill_manual(values=cols)+
  labs(x="APOE genotype", y="Allele score") + 
  scale_x_discrete( labels=c(expression(paste(epsilon,"22")),expression(paste(epsilon,"32")),expression(paste(epsilon,"33")),expression(paste(epsilon,"42")),expression(paste(epsilon,"43")),expression(paste(epsilon,"44")))) + 
  scale_y_discrete( labels=c("0-25%","25-50%","50-75%","75-100%")) + 
  theme(axis.ticks = element_blank(), panel.background = element_blank(),legend.position = "none",      axis.title=element_text(size=12,face="bold")) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0 , b = 0, l = 0))) +
  theme(axis.text=element_text(size=8,face="italic"), strip.background = element_blank()) +
  facet_grid(age_gr ~ dm + smoker , labeller = labeller(age_gr = age_gr.labs, smoker = smoker.labs, dm = dm.labs)) +
  theme(strip.text.x = element_text( size = 10, face = "bold"),
        strip.text.y = element_text( size = 10, face = "bold")) +
  ggtitle(titlename) +
  coord_fixed()


plotname <- paste0("./Results/",cohortname,"_EADB2022_Riskplot_10y_vard_sex",i,"_edu",j,".pdf")

ggsave(plotname,device="pdf",width=20,dpi=300,units="cm")

}
}


```

```{r alldem risk charts}

# All dementia risk charts

alldem_df <- alldem_df %>% mutate(apoe_s = case_when(
                  apoe == "e22" ~ 1,
                  apoe == "e32" ~ 2,
                  apoe == "e33" ~ 3,
                  apoe == "e42" ~ 4,
                  apoe == "e43" ~ 5,
                  apoe == "e44" ~ 6)
                )

alldem_df$apoe_s<- as.factor(alldem_df$apoe_s)

alldem_df <- alldem_df %>% mutate(col = case_when(
                  alldem_pred_10 < 0.01 ~ "0",
                  alldem_pred_10 >= 0.01 & alldem_pred_10 < 0.05 ~ "1",
                  alldem_pred_10 >= 0.05 & alldem_pred_10 < 0.10 ~ "2",
                  alldem_pred_10 >= 0.10 & alldem_pred_10 < 0.15 ~ "3",
                  alldem_pred_10 >= 0.15 & alldem_pred_10 < 0.20 ~ "4",
                  alldem_pred_10 >= 0.20 & alldem_pred_10 < 0.30 ~ "5",
                  alldem_pred_10 >= 0.30 ~ "6")
                )


summary(alldem_df$age_gr)
alldem_df$age_gr <- factor(alldem_df$age_gr, levels = c(4, 3, 2, 1,0))


for (i in 0:1) {
  for (j in 0:1) {

pd_alldem <- alldem_df[sex==i & educate==j & age_gr != 0,]

titlename <- paste0(cohortname," - All dementia 10 year absolute risk - sex=" ,i, " edu=" ,j)

ggplot(pd_alldem,aes(x = apoe_s, y = was2022_gr, fill = col))  +
  geom_tile() +
  geom_text(aes(label = lapply(alldem_pred_10*100.0, as.integer),fontface = "bold")) +
  scale_fill_manual(values=cols)+
  labs(x="APOE genotype", y="Allele score") + 
  scale_x_discrete( labels=c(expression(paste(epsilon,"22")),expression(paste(epsilon,"32")),expression(paste(epsilon,"33")),expression(paste(epsilon,"42")),expression(paste(epsilon,"43")),expression(paste(epsilon,"44")))) + 
  scale_y_discrete( labels=c("0-25%","25-50%","50-75%","75-100%")) + 
  theme(axis.ticks = element_blank(), panel.background = element_blank(),legend.position = "none",      axis.title=element_text(size=12,face="bold")) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0 , b = 0, l = 0))) +
  theme(axis.text=element_text(size=7,face="italic"), strip.background = element_blank()) +
  facet_grid(age_gr ~ dm + smoker , labeller = labeller(age_gr = age_gr.labs, smoker = smoker.labs, dm = dm.labs)) +
  theme(strip.text.x = element_text( size = 10, face = "bold"),
        strip.text.y = element_text( size = 10, face = "bold")) +
  ggtitle(titlename) +
  coord_fixed()

plotname <- paste0("./Results/",cohortname,"_EADB2022_Riskplot_10y_alldem_sex",i,"_edu",j,".pdf")

ggsave(plotname,device="pdf",width=20,dpi=300,units="cm")

}
}

rm(age_gr.labs,cols,educate.labs,i,j,smoker.labs,plotname,titlename)

```

## Generate 10 year risk-ratios
```{r generate risk ratios}
# Women, no diabetes, never smoking, e33, 3rd allele_gr, age_gr 70-80, Normal education is selected as reference

# Alzheimer
ref_m <- filter(alz_df, sex==0 & dm==0 & smoker == 0 &  educate == 0 & apoe =="e33" & was2022_gr == 2 & age_gr == 3)
ref <- as.double(ref_m$alz_pred_10)


alz_df <-alz_df %>% mutate( alz_rr_10 = alz_pred_10 / ref)

# Vascular
ref_m <- filter(vas_df, sex==0 & dm==0 & smoker == 0 &  educate == 0 & apoe =="e33" & was2022_gr == 2 & age_gr == 3)
ref <- as.double(ref_m$vas_pred_10)


vas_df <-vas_df %>% mutate( vas_rr_10 = vas_pred_10 / ref)

# vascular related 
ref_m <- filter(vard_df, sex==0 & dm==0 & smoker == 0 &  educate == 0 & apoe =="e33" & was2022_gr == 2 & age_gr == 3)
ref <- as.double(ref_m$vard_pred_10)


vard_df <- vard_df %>% mutate( vard_rr_10 = vard_pred_10 / ref)

# All_dementia

ref_m <- filter(alldem_df, sex==0 & dm==0 & smoker == 0 &  educate == 0 & apoe =="e33" & was2022_gr == 2 & age_gr == 3)
ref <- as.double(ref_m$alldem_pred_10)


alldem_df <-alldem_df %>% mutate( alldem_rr_10 = alldem_pred_10 / ref)


```

# Save risk estimates and workspace

```{r save risk estimates and workspace}

# Saving risk prediction dataframe and FG-models for each endpoint and furthermore the cohort-maf in a R-datafile

save(alz_df,alz_fgr_was,alz_cox_was,vas_df,vas_fgr_was,vas_cox_was,vard_df,vard_fgr_was,vard_cox_was,alldem_df,alldem_fgr_was,alldem_cox_was,gsmodel2022, file = paste0("./Results/Risk_estimates_10y_EADB2022_",cohortname,".Rdata"))

save.image(file = paste0("./data/Workspace_10y_EADB2022_",cohortname,".Rdata"))


```





